A Monte Carlo Synthetic Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-05-26 doc/jcomp2012/jcomp2012.tex
Adding adjoint solver justification to jcomp2012 paper

>>>>

\subsection{Justification of Adjoint Error Estimation}

The MCSA method defined in Eq.(~\ref{eq:general-MCSA}) uses the
adjoint method in Eq.~(\ref{eq:MCSA-residual_solve}) to estimate the
error instead of the direct method. To demonstrate the effectiveness
of the adjoint method over the direct method, we choose the
two-dimensional time-dependent heat equation as a simple model
problem

\begin{equation}
  \frac{\partial u}{\partial t} = \alpha \nabla^2 u
  \label{eq:heat_equation}
\end{equation}

where $\alpha$ is a constant coefficient. For all comparisons, a
single time step is computed with implicit Euler time integration. The
Laplacian is differenced on a square Cartesian grid with a
second-order five point stencil

\begin{equation}
  \nabla^2_5 = \frac{1}{h^2}[u_{i-1,j} + u_{i+1,j} + u_{i,j-1} +
    u_{i,j+1} - 4 u_{i,j}]
  \label{eq:five_point_stencil}
\end{equation}

and a fourth-order nine point stencil both assuming a grid size of $h$
in both the $i$ and $j$ directions.

\begin{equation}
  \nabla^2_9 = \frac{1}{6h^2}[4 u_{i-1,j} + 4 u_{i+1,j} + 4 u_{i,j-1}
    + 4 u_{i,j+1} + u_{i-1,j-1} + u_{i-1,j+1} + u_{i+1,j-1} +
    u_{i+1,j+1} - 20 u_{i,j}]
  \label{eq:nine_point_stencil}
\end{equation}

For a single time step solution, we then have the following linear
system to be solved with the MCSA method.

\begin{equation}
  A u^{n+1} = u^n
  \label{eq:heat_eq_lin_sys}
\end{equation}

Both the stencils will be used to vary the size and density of the
sparse linear system in Eq.~(\ref{eq:heat_eq_lin_sys}).

A timing study and convergence study is used to demonstrate the
effectiveness of the adjoint method as compared to the direct
method. To assess both the CPU time and number of iterations required
to converge to a solution, a problem of constant grid size and
$\alpha$ was used with varying values of problem size, fixing the
spectral radius of the system at a constant value for each
variation. Both the five point and nine point stencils were used with
both the direct and adjoint solvers. For each case, 50 histories were
simulated per cycle with a weight cutoff of $1.0E-4$ for the
stochastic solver and a convergence tolerance of $1.0E-8$ for the MCSA
solver. All computations were completed on a 2.66 GHz Intel Core i7
machine with 8 GB 1067 MHz DDR3 memory.

Fig.~\ref{fig:hmcsa_cpu_time} gives the CPU time needed for each case
to converge in seconds and Fig.~\ref{fig:hmcsa_iterations} gives the
number of iterations needed for each case to converge. We see clearly
in Fig.~\ref{fig:hmcsa_cpu_time} that the using the adjoint solver
with MCSA results in several orders of magnitude speedup over the
direct solver while the number of iterations required to converge is
of a similar scale. We expect this for several reasons. First, with an
equivalent number of histories specified for both solvers and a system
of size $N$, the direct solver will compute $N$ times the number of
histories random walks while the adjoint solver will only compute the
number of histories specified. Initially, this may not seem like a
fair comparison, however, we see from Fig.~\ref{fig:hmcsa_cpu_time}
that the number of iterations requried to converge is approximately
the same. This is necessary in the direct method to ensure a
contribution from each state as the random walk sequence will only
contribute to the starting state. For the adjoint method, because the
random sequence contributes to the state in which it currently
resides, fewer histories are neccessary to compute contributions from
all the nonzero parts of the source. Therefore, although the total
number of histories per cycle is in fact larger for the direct method,
the solution computed at each iteration is no better than that
computed by the adjoint method.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_CPU_Time.eps}
  \caption{\sl CPU Time (s) to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. A significant CPU time speedup
    is noted with the adjoint method due to the reduced number of
    histories.}
  \label{fig:hmcsa_cpu_time}
\end{figure}

\begin{figure}[htbp!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Iterations.eps}
  \caption{\sl Iterations to converge vs. Problem Size (N for an NxN
    problem). Both the adjoint and direct solvers are used with the
    five point and nine point stencils. Both methods give similar
    iteration behavior.}
  \label{fig:hmcsa_iterations}
\end{figure}

Although CPU time is shown to be greatly decreased by using the
adjoint solver in Fig.~\ref{fig:hmcsa_cpu_time}, the iteration
behavior was approximately the same. To assess the convergence
properties of MCSA using each solver and stencil, the infinity norm of
the residual computed in Eq.~(\ref{eq:general-MCSA}) was computed at
each iteration for a fixed problem
size. Fig.~\ref{fig:hmcsa_convergence} gives the results of these
computations. First, it is worthy to note on the semilog plot that we
are indeed achieving the expected exponential convergence from
MCSA. Second, we note that both the direct and adjoint methods have
approximately equivalent convergence behavior.

\begin{figure}[htpb!]
  \centering
  \includegraphics[width=5in,clip]{Adjoint_Direct_Convergence.eps}
  \caption{\sl Infinity norm of the solution residual vs. iteration
    number for a problem of fixed size. Both the adjoint and direct
    solvers are used with the five point and nine point stencils. Both
    methods give similar iteration behavior.}
  \label{fig:hmcsa_convergence}
\end{figure}
